Project Diary

### 2014-05-08 thesis/price_models.tex
This is the version to be used for job searching.
Change the section about Nordea 15min in 1st quarter 2012.

\chapter{Return Models}\label{chp:PriceModels}
In this chapter we review some of the discrete-time return models and
fit them to intraday returns. The intention is to assess the
applicability of these models and to understand their statistical
properties. The implications of these properties on the returns'
correlation and covariance matrix are studied in chapters
\ref{chp:Gaussian} and \ref{chp:CrossCorrelationFat}.

\S\ref{sec:Garch_model} reviews the GARCH model using the Nordea
15-minute returns as an example. \S\ref{sec:SLV_model} examines the
stochastic log-volatility model and discusses the possibility of
describing the residuals of the log-volotility series using the
Johnson Su distribution. The same Nordea return series is fitted as an
example. \S\ref{sec:XieCalc} derives the unconditional distribution
function of the stochastic log-volotility model and compares the
analytic results to the 30-minute returns of a number of Swedish
stocks.

\section{Gaussian Distribution}
The justification of modeling return series as independent,
identically distributed Gaussian variates comes from imagining the
price process $S(t)$ as a Brownian motion whose increment $\sigma
dw_t$ at each time step is independent and scales as $\sigma
\sqrt{dt}$. Then, by adding a drift term $\mu dt$ that represents some
deterministic trend in the price process, one can express the
price $S(t)$ as a stochastic differential equation:
\begin{eqnarray*}
  dS &=& S\mu dt + S\sigma dw_t \\
  d(\ln S) &=& (\mu - \frac{1}{2} \sigma^2)dt + \sigma dw_t \\
\end{eqnarray*}
It follows from this equation that $\ln S(t) - \ln S_0$, where $S_0$
is the price at time 0, has Gaussian distribution with mean $(\mu -
\sigma^2/2)t$ and variance $\sigma^2 t$. Therefore, in a discrete-time
model, where the length of each time step $\Delta t$ is fixed, the
return $r_t = \ln S(t + \Delta t) - \ln S(t)$ is assumed to be
Gaussian distributed and have mean and variance that are functions of
$\Delta t$.

@@

\section{GARCH models} \label{sec:Garch_model}
``GARCH'' is the acronym for ``Generalized Autoregressive Conditional
Heteroscedasticity''. A GARCH(p, q) model is defined by the following
equation system:
\begin{equation}
  \label{eq:garch_def}
  \begin{aligned}
    r_t &= \mu + \epsilon_t \\
    \sigma_t^2 &= \alpha_0 + \sum_{i=1}^q \alpha_i \epsilon_{t-i}^2 +
    \sum_{i=1}^p \beta_i \sigma_{t-i}^2
  \end{aligned}
\end{equation}
where $\epsilon_t$ is a random variable of a prescribed distribution,
the most common ones being Gaussian, Student's t, Asymmetric Student's
t, etc; $\sigma_t^2$ is the variance of $\epsilon_t$
\cite{Bollerslev86}. The autocorrelation function of $\epsilon_t^2$ is
given by (c.f. \cite{Bollerslev87})
\begin{eqnarray*}
\rho_n &=& \sum_{i=1} ^{p \vee q} (\alpha_i + \beta_i) \rho_{n-i}
\text{ for $n > p$}
\end{eqnarray*}
where $\alpha_i$ with $i > q$ and $\beta_i$ with $i > p$ are taken as
zeros. $p \vee q$ denotes the maximum of p and q. From these
equations, it is clear that the partial autocorrelation function cuts
off at $\max(p, q)$.

To assess the applicability of GARCH models at the face of intra-day
returns, we study the {\it Nordea Bank 15min} returns sampled during
the period 2012/01/16 - 2012/04/20. Firstly we plot the autocorrelation
function (ACF) of $\epsilon_t^2$ from this sample, as shown
in figure \ref{fig:nordea_15min_acf}.

At the absence of GARCH effects, the ACF is expected to have an
asymptotic Gaussian distribution with mean 0 and variance 1/T
\cite{Bollerslev86, Bollerslev87}, which is clearly not the case in
figure \ref{fig:nordea_15min_acf}: The first 5 autocorrelations have
comparable sizes and do not fall off as in a Gaussian
scheme. Moreover, figure \ref{fig:nordea_15min_vlt_acf} shows 
even more clearly that the conditional variances of the series are
correlated. These observations suggest a GARCH(p, q) model can be
appropriate.

\begin{figure}[htb!]
  \centering
  \subfigure[]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_acf.pdf}
    \label{fig:nordea_15min_acf}
  }
  \subfigure[]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_vlt_acf.pdf}
    \label{fig:nordea_15min_vlt_acf}
  }
  \caption{\small \it Auto-correlations among Nordea Bank 15min
    returns. \ref{fig:nordea_15min_acf}: Auto-correlations (ACF) of the
    squared returns; \ref{fig:nordea_15min_vlt_acf}: Auto-correlations
    (ACF) of the conditional variances.}
\end{figure}

According to Andersen and Bollerslev et al \cite{Andersen03},
$\epsilon_t$ can be decomposed as $\epsilon_t = \sigma_t z_t$, where
$z_t \sim N(0, 1)$ and $\sigma_t$ can be estimated using realized
volatility. To find out whether this is true for the aforementioned
series we look at the probability plot of $\epsilon_t/\sigma_t$
(figure \ref{fig:nordea_bank_15min_z_prob}), which is consistent with
a standard Gaussian distribution except for a few outliers at the two
ends. Hence the fat tails of the probability density function can be
entirely accounted for by GARCH effects, implying a Gaussian
conditional distribution is sufficient for $z_t$.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=80 258 104
    220]{../pics/nordea_bank_15min_z_prob.pdf}
  \caption{\small \it{Probability plot of $z_t = \epsilon_t /
      \sigma_t$. $\epsilon_t$ are derived from Nordea Bank 15min
      returns while $\sigma_t$ are realized volatilities calculated
      using 30s returns within each 15min interval. Horizontal axis:
      $z_t$}; Vertical axis: cummulative probability function (CDF) of
    $z_t$, arranged on such a scale that the CDF of the standard
    Gaussian is a straight line.}
  \label{fig:nordea_bank_15min_z_prob}
\end{figure}

To determine the orders p and q of the GARCH model, we again check
the partial autocorrelation function of $\epsilon_t^2$. As just
mentioned, the first 5 partial autocorrelations are significant. Thus
we can tentatively estimate a GARCH(1, 5) model. Then by applying the
Lagrange multiplier test and accordingly removing insignificant
parameters, we obtain a GARCH(1, 3) model
\ref{tab:nordea_15min_garch}:
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    parameter & value & std. error & t statistic \\
    \hline
     $\alpha_0$ &   1.05668e-06 &   3.56871e-07 &    2.96096\\
     \hline
     $\beta_1$  &     0.687212  &   0.0299314   &    22.9596\\
     \hline
     $\alpha_1$ &      0.128614 &   0.0268144   &    4.79643\\
     \hline
     $\alpha_3$ &    0.0568022  &   0.0301503   &    1.88397 \\
     \hline
  \end{tabular}
  \caption{\small \it GARCH(1, 3) model of Nordea Bank 15min returns}
  \label{tab:nordea_15min_garch}
\end{table}

\section{ARIMA Log-volotility Models}\label{sec:SLV_model}
With the availability of transaction data, estimating conditional
volatility using returns sampled at a higher frequency (realized
volatility) gives superior acuracy and reliability. However, at which
frequency the time series should be sampled in order to give an
unbiased and consistent estimate of the volatility is not a trivial
question. Naively one would argue by the doctrine of statistics that
the often the series is sampled, the better the estimate. This is also
well supported in theory, since the quadratic variation which is the
theoretial variance of the time series (see \cite{Andersen03}, theorem
1 and corollary 1), is approximated as (see \cite{Protter05}, ch.II,
\S6, theorem 22):
\[
\var(p_{t+h} - p_t) = [r, r]_{t+h} - [r, r]_t = \lim_{n \to \infty}
\sum_{T^i \in \pi_n} (p^{T_{i+1}} - p^{T_i})^2
\]
where $p_t$ is the log price process and $r_t = p_t - p_0$. $\pi_n$
is a refining partition of the interval $[t, t+h]$.

However, due to the noise introduced by market micro-structure, there
is actually an optimal sampling frequency relative to $h$. While the
method to determine this optimal frequency is a subject of debate (see
for example \cite{Sahalia05}), it is not difficult to find a fairly
satisfactory frequency in practice.

Andersen and Bollerslev et al proved in \cite{Andersen03} that, with
the conditional volatility well approximated by the realized
volatility, $r_{t+h, h} = p_{t+h} - p_t$ is governed by a Gaussian
distribution (see theorem 2 of \cite{Andersen03}):
\[
r_{t+h, h} | \sigma(\mu_{t+s}, \sigma_{t+s})_{s \in [0, h]} \sim
N(\int_{0}^h \mu_{t+s} ds, \int_{0}^h \sigma_{t+s}^2 ds)
\]

Thus one can try a few frequencies and compare the distribution of
$r_{t+h, h}/\hat{\sigma}_{t+h, h}$ with the standard Gaussian. If the
two match, $\hat{\sigma}_{t+h, h}$ is a good estimate of $\sigma_{t+h,
  h}$, the volatility of $r_{t+h, h}$.

For the Nordea Bank 15-minute returns, $h = 15\text{min}$, and a
reasonally good sampling frequency is 1/30sec, as confirmed by the
probability plot \ref{fig:nordea_bank_15min_z_prob} as well as by the
autocorrelation functions of $r_{t+h, h} / \sigma_{t+h, h}$ and of
$(r_{t+h, h} / \sigma_{t+h, h})^2$ which are shown in
\ref{fig:nordea_15min_quotient_acf} and
\ref{fig:nordea_15min_quotient_squared_acf} respectively.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of $r_{t+h, h} / \sigma_{t+h, h}$]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_quotient_acf.pdf}
    \label{fig:nordea_15min_quotient_acf}
  }
  \subfigure[ACF of $(r_{t+h, h} / \sigma_{t+h, h})^2$]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_quotient_squared_acf.pdf}
    \label{fig:nordea_15min_quotient_squared_acf}
  }
  \caption{\small \it Nordea 15min $r_t/\sigma_t$ and
    $(r_t/\sigma_t)^2$ ACF.}
\end{figure}

We see in figure \ref{fig:nordea_15min_quotient_acf} and
\ref{fig:nordea_15min_quotient_squared_acf} there is essentially no
autocorrelation in the series $r_{t+h, h} / \sigma_{t+h, h}$ and
$(r_{t+h, h} / \sigma_{t+h, h})^2$, supporting the assumption that the
series is i.i.d Gaussian distributed. As a result, the problem of
modeling $r_{t+h, h}$ boils down to modeling $\sigma_{t+h,
  h}$. Andersen and Bollerslev et al reported that, for the exchange
rates between Deutch mark, yen and dollar, $\ln \sigma_{t+h, h}$ is
gaussian distributed \cite{Andersen03}. This is, however, not the case
for our series in question. In fact, in our case, $\ln \sigma_{t+h,
  h}$ is right skewed (skewness 0.3342) and leptokurtic (kurtosis
6.1006). See figure \ref{fig:nordea_15min_logvol_prob}
\begin{figure}[htb!]
  \centering
  %\vspace{-15mm}
  \includegraphics[scale=0.5, clip=true, trim=80 223 107
  4]{../pics/nordea_15min_logvol_prob.pdf}
  \caption{\small \it Probability plot of Nordea 15min
    $\ln\sigma_{t+h, h}$ unconditional distribution}
  \label{fig:nordea_15min_logvol_prob}
\end{figure}

Moreover, the series of $\ln\sigma_{t+h, h}$ shows long-lasting and
periodic autocorrelations with an apparent period of 33 (see figure
\ref{fig:nordea_15min_logvol_acf}). This suggests the series may be
described by an integrated seasonal ARMA\footnote{ARMA ---
  Autoregressive Moving Average} model. But before we set out
to specify and estimate this model of the logvolatility
$\ln\sigma_{t+h, h}$, it is useful to first write down the model of
the returns $r_t$:
\begin{eqnarray}
   r_{t+h, h} &=& \ln p_{t+h} - \ln p_t \nonumber \\
   r_{t+h, h} &=& \mu + \sigma_{t+h, h} b_t \label{eq:SLV_spec}
\end{eqnarray}
where $p_t$ is the price of the asset at time $t$ and $b_t \sim N(0,
1)$. For simplicity, in the rest of this chapter we shall just write
$t$ for the subscript ``t+h, h'', since the time interval $h$ is fixed
for each time series and is a known constant. In the case of the
Nordea Bank 15-minute returns, h = 15 minutes.

As has just been said, $\ln \sigma_t$ for the series in question has
long-lasting auto-correlations (often referred to as ``long
memory''). This suggests simplifying the series by differencing
\cite{BoxJenkins94}:
\[
w_t = (1-B)(1-B^s)\ln\sigma_t
\]
where $B$ is the back-shift operator and $s=33$ is the seasonality.

The autocorrelation function of the differenced process $w_t$, as shown in
figure \ref{fig:nordea_15min_w_acf}, clearly points to a seasonal
moving-average model: There are only 4 non-zero autocorrelations in
the plot, located at lags 1, 32, 33, 34, respectively; furthermore,
the two at 32 and 34 are approximately equal. Thus we can write down
the model as
\begin{eqnarray}
  w_t &=& (1 - \theta B)(1 - \Theta B^s) y_t \label{eq:nordea_w}
\end{eqnarray}
where $\theta$ and $\Theta$ are parameters to be determined and $y_t$
is a noise process with constant variance $\sigma_y^2$ and mean
0. $y_t$ is often refered to as the residuals.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of $\ln\sigma_{t+h,h}$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_logvol_acf.pdf}
    \label{fig:nordea_15min_logvol_acf}
  }
  \subfigure[ACF of $w_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_w_acf.pdf}
    \label{fig:nordea_15min_w_acf}
  }
  \caption{\small \it Nordea 15min $\ln\sigma_{t+h, h}$ and $w_t$
    autocorrelations}
\end{figure}

@@

With an estimate of $\theta$ and $\Theta$, one can then infer the
noise process i.e. the residuals $y_t$:
\begin{equation}
  \label{eq:infer_y}
  y_t = w_t + \theta y_{t-1} + \Theta y_{t-s} - \theta \Theta y_{t-s-1}
\end{equation}
where we substitute $y_t\;(t \leq 0)$ with their unconditional
expectation 0.
% By inverting the seasonal moving average model of $w_t$ we can then
% infer the noise process $y_t$ \footnotemark, i.e. the residuals:
% \begin{eqnarray*}
%   y_t &=& {1 \over 1 - \theta B}{1 \over 1 - \Theta B^s} w_t \\
%   &=& \sum_{i=0}^{\infty}\sum_{j=0}^{\infty} \theta^i \Theta^j B^{i+j}
%   w_t \\
%   &=& w_t + \sum_{k=1}^{\infty} \sum_{j=0}^{\bot(k/s)} \theta^{k-sj} \Theta^j
%   w_{t-k}
% \end{eqnarray*}
% where $\bot(\cdot)$ denotes the floor function, i.e. the largest
% integer smaller than its argument.
% \footnotetext{
%   The autoregressive process resulting from the inversion is inifinite
%   in extent. Hence one must choose a cut-off value for k. For the
%   present case of Nordea 15min log-volatility, I choose the value
%   $k=sj$ for which $|\Theta|^j < 0.02$. This value of k, call it
%   $k_{\top}$, is around 550. To avoid data reduction of this amount,
%   we add to the beginning of $\{w_t\}_{t=1}^n$ $k_{\top}$ zeros
%   corresponding to t=0, -1, ..., -$k_{\top}$+1. Recall $\E(w_t) = \E(y_t
%   - \theta y_{t-1} - \Theta y_{t-s} + \theta \Theta y_{t-s-1}) = 0$.
% }

In order to forecst the $w_t$ process, and hence the return process
itself, we must also know the distribution of $y_t$. Moreover, to
properly estimate the parameters of the model in the sense of maximum
likelihood, we are also in need of the distribution of $y_t$.

If $y_t$ is indeed i.i.d random variables as we assumed, we only need
to work out its unconditional distribution. This is shown in figure
\ref{fig:nordea_15min_y_qq} and \ref{fig:nordea_15min_y_acf}.
\begin{figure}[htb!]
  \centering
  \subfigure[QQ-plot of $y_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    210]{../pics/nordea_15min_y_qq.pdf}
    \label{fig:nordea_15min_y_qq}
  }
  \subfigure[ACF of $y_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    210]{../pics/nordea_15min_y_acf.pdf}
    \label{fig:nordea_15min_y_acf}
  }
  \caption{\small \it Nordea 15min $y_t$ QQ-plot and autocorrelations}
\end{figure}
In addition, the measured moments of $y_t$ are shown in table
\ref{tab:nordea_15min_y_moments}:
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    mean & variance & skewness & kurtosis \\
    \hline
    0.0012 & 0.0935 & 0.2988 & 6.8691 \\
    \hline
  \end{tabular}
  \caption{\small \it Nordea 15min $y_t$ moments}
  \label{tab:nordea_15min_y_moments}
\end{table}

If we ignore the weak autocorrelations in this crude estimate of
$y_t$ and assume $y_t$ to be i.i.d random variables governed
by a skewed and leptokertic distribution, the parameters of this
distribution together with $\theta$ and $\Theta$ can be determined via
maximum likelihood estimate (MLE).

It turns out that $y_t$ can be well described by a Johnson Su
distribution \cite{Shang2004}:
\[
  y_t = \xi + \lambda\sinh{z_t - \gamma \over \delta}
\]
where $\gamma, \delta, \lambda, \xi$ are parameters to be determined
and $z_t \sim N(0, 1)$. The goodness of fitting is demonstrated in
figure \ref{fig:nordea_15min_y_js_fit} by the
empirical cummulative distribution function in comparison to the
theoretical one.
\begin{figure}[htb!]
  %\vspace{-18mm}
  \centering
    \includegraphics[scale=0.5, clip=true, trim=92 229 116
    133]{../pics/nordea_15min_y_js_fit.pdf}
    \caption{\small \it Nordea 15min $y_t$ fitted to a Johnson Su
      distribution. Horizontal: values of $y_t$, denoted x; Vertical:
      $\ln\left(P(y_t < x)\right)$.}
    \label{fig:nordea_15min_y_js_fit}
\end{figure}

>>>>

\chapter{Return Models}\label{chp:PriceModels}
In this chapter we review some of the discrete-time return models and
fit them to intraday returns. The intention is to compare these models
in terms of forecast accuracy and to understand their statistical
properties. In the following we first describe these models briefly,
then in

% \S\ref{sec:Garch_model} reviews the GARCH model using the Nordea
% 15-minute returns as an example. \S\ref{sec:SLV_model} examines the
% stochastic log-volatility model and discusses the possibility of
% describing the residuals of the log-volotility series using the
% Johnson Su distribution. The same Nordea return series is fitted as an
% example. \S\ref{sec:XieCalc} derives the unconditional distribution
% function of the stochastic log-volotility model and compares the
% analytic results to the 30-minute returns of a number of Swedish
% stocks.

\begin{enumerate}
\item{\bf Gaussian Distribution}

The justification of modeling return series as independent,
identically distributed Gaussian variates comes from imagining the
price process $S(t)$ as a Brownian motion whose increment $\sigma
dw_t$ at each time step is independent and scales as $\sigma
\sqrt{dt}$. Then, by adding a drift term $\mu dt$ that represents some
deterministic trend in the price process, one can express the
price $S(t)$ as a stochastic differential equation:
\begin{eqnarray*}
  dS &=& S\mu dt + S\sigma dw_t \\
  d(\ln S) &=& (\mu - \frac{1}{2} \sigma^2)dt + \sigma dw_t \\
\end{eqnarray*}
It follows from this equation that $\ln S(t) - \ln S_0$, where $S_0$
is the price at time 0, has Gaussian distribution with mean $(\mu -
\sigma^2/2)t$ and variance $\sigma^2 t$. Therefore, in a discrete-time
model, where the length of each time step $\Delta t$ is fixed, the
return $r_t = \ln S(t + \Delta t) - \ln S(t)$ is assumed to be
Gaussian distributed and have mean and variance that are functions of
$\Delta t$.

@@

\item{\bf GARCH models}
``GARCH'' is the acronym for ``Generalized Autoregressive Conditional
Heteroscedasticity''. A GARCH(p, q) model is defined by the following
equation system:
\begin{equation}
  \label{eq:garch_def}
  \begin{aligned}
    r_t &= \mu + \epsilon_t \\
    \sigma_t^2 &= \alpha_0 + \sum_{i=1}^q \alpha_i \epsilon_{t-i}^2 +
    \sum_{i=1}^p \beta_i \sigma_{t-i}^2
  \end{aligned}
\end{equation}
where $\epsilon_t$ is termed the innovation of the return $r_t$ and is
a random variable of a prescribed distribution, the most common ones
being Gaussian, Student's t, Asymmetric Student's t, etc; $\sigma_t^2$
is the variance of $\epsilon_t$ \cite{Bollerslev86}. The
autocorrelation function of $\epsilon_t^2$ is given by
(c.f. \cite{Bollerslev87})
\begin{eqnarray*}
\rho_n &=& \sum_{i=1} ^{p \vee q} (\alpha_i + \beta_i) \rho_{n-i}
\text{ for $n > p$}
\end{eqnarray*}
where $\alpha_i$ with $i > q$ and $\beta_i$ with $i > p$ are taken as
zeros. $p \vee q$ denotes the maximum of p and q. From these
equations, it is clear that the partial autocorrelation function cuts
off at $\max(p, q)$.

\item {\bf Stochastic Volatility Models}\label{sec:SLV_model}
For the purpose of intraday returns, we specify the stochastic
volatility model as
\begin{eqnarray}
   r_{t, t-h} &=& \ln p_{t} - \ln p_{t-h} \nonumber \\
   r_{t, t-h} &=& \mu + \sigma_{t, t-h} b_t \label{eq:SLV_spec}
\end{eqnarray}
where $p_t$ is the price of the asset at time $t$; $b_t \sim N(0,
1)$; $r_{t, t-h}$ is the return in the time interval $[t-h, t]$.
For simplicity, in the rest of this chapter we shall just write
$t$ for the subscript ``t, t-h'', since the time interval $h$ is fixed
for each time series and is a known constant.

Andersen and Bollerslev et al proved the following in
\cite{Andersen03} (theorem 2):
\begin{equation}
  \label{eq:normal_r}
  r_t|\mathcal{F}_{t-h} \sim N(\int_{0}^h \mu_{t-h+s} ds, \int_{0}^h
  \sigma_{t-h+s}^2 ds)
\end{equation}
In plain words, given all the information up to time $t-h$, the
distribution of $r_t$ is Gaussian with integrated mean and variance.
The variance of this conditional distribution, $\int_{0}^h
\sigma_{t-h+s}^2 ds$, can be approximated by \cite{Protter05,
  Andersen03}:
\begin{equation}
  \label{eq:rv_def}
  \int_{0}^h \sigma_{t-h+s}^2 ds = \sum_{k=1}^n \left(
    p_{t-h+kh/n} - p_{t-h+(k-1)h/n} \right)^2
\end{equation}
where $n$ is a chosen constant. The square root of the right hand side
of equation \ref{eq:rv_def} is the realized volatility, call it
$\hat{\sigma}_t$.

With the availability of transaction data, estimating conditional
volatility using returns sampled at a higher frequency (realized
volatility) gives superior accuracy and reliability. However, at which
frequency the time series should be sampled (the choice of $n$) in
order to give an unbiased and consistent estimate of the volatility is
not a trivial question. Naively one would believe that the often the
series is sampled, the better the estimate, but in fact, due to noise
introduced by market micro-structure, there is an optimal sampling
frequency, depending on $h$. While the method to determine this
optimal frequency is a subject of debate (see for example
\cite{Sahalia05}), it is not difficult to find a fairly satisfactory
frequency in practice:

Keeping equation \ref{eq:normal_r} in mind, one can simply try a few
frequencies and compare the distribution of $(r_t -
\E(r_t))/\hat{\sigma}_t$ with the standard Gaussian. If the two match,
$\hat{\sigma}_t$ is a good approximation of
\[
\sigma_t = \left(\int_{0}^h \sigma_{t-h+s}^2 ds \right)^{1/2}
\]

\end{enumerate}

To compare GARCH and stochastic log-volatility models at the face of
intraday returns, we study the {\it Nordea Bank} 15-minute returns
during the period 2012/01/16 - 2012/04/20 in \S\ref{chp:nordea_15min},
and {\it Volvo B} 30-minute returns during 2013/10/10 - 2014/04/04.

\section{Case Study: Nordea in 1st quarter 2012}
\label{chp:nordea_15min}
In this section we investigate the Nordea 15-minute returns sampled
during the period 2012/01/16 - 2012/04/20. In total, these amount to
2022 returns. We use the first 80\% (1617) for model estimation and  the
remaining 20\% (405) for comparing with model forcasts. In section
\ref{sec:nordea_15min_garch} we study the series with a GARCH model 
and in section \ref{sec:nordea_15min_arima} we study it with a
stochastic volatility model.

\subsection{GARCH Model}\label{sec:nordea_15min_garch}
First of all, to check for GARCH effects, we plot the auto-correlation
function (ACF) of the squared returns. This is shown in figure
\ref{fig:nordea_15min_acf}. At the absence of GARCH effects, the ACF
is expected to have an asymptotic Gaussian distribution with mean 0
and variance 1/T \cite{Bollerslev86, Bollerslev87}, which is clearly
not the case in figure \ref{fig:nordea_15min_acf} --- the first 5
autocorrelations have comparable sizes and do not fall off as in a
Gaussian scheme. Moreover, figure \ref{fig:nordea_15min_vlt_acf} shows
even more clearly that the conditional variances of the series are
correlated. These observations suggest a GARCH(p, q) model can be
appropriate.

Starting with a GARCH(1,1) model and taking advantage of the knowledge
that the log-volatility $\ln \sigma_t$ has seasonality $s=33$, we fit
to the return series a GARCH(33, 33) model, limiting to lags 1 and 33
for both ARCH and GARCH parameters.
\begin{eqnarray*}
  r_t &=& \mu + €_t \\
  €_t &=& \sigma_t z_t \\
  \sigma^2_t &=& \alpha_0 + \alpha_1 €^2_{t-1} + \alpha_s €^2_{t-s} +
  \beta_1 \sigma^2_{t-1} + \beta_s \sigma^2_{t-s}
\end{eqnarray*}
Via maximum likelihood estimation, parameter values listed in table
\ref{tab:nordea_15min_garch} are obtained.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    Parameter & $\alpha_0$ & $\alpha_1$ & $\alpha_s$ & $\beta_1$ &
    $\beta_s$ \\
    \hline
    Value & $4.7833 \times 10^{-7}$ & 0.1600 & 0.0667 & 0.6846 &
    0.0342 \\
    \hline
  \end{tabular}
  \caption{\small \it GARCH model parameters}
  \label{tab:nordea_15min_garch}
\end{table}

\subsection{Stochastic Volatility Model}\label{sec:nordea_15min_arima}
For the Nordea Bank 15-minute returns under consideration, it can be
verified that the square root of the sum of squared 30-second returns
makes a good proxy for the volatility. This can be seen from the
probability plot of $z_t = (r_t - \E(r_t))/\hat{\sigma}_t$ (figure
\ref{fig:nordea_bank_15min_z_prob}), i.e. the quotient of the 
15-minute returns over the volatility proxy.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.4, clip=true, trim=80 258 104
    220]{../pics/nordea_bank_15min_z_prob.pdf}
  \caption{\small \it{Probability plot of $z_t =
      (\epsilon_t-\E(\epsilon_t))/\sigma_t$. $\epsilon_t$ are derived
      from Nordea Bank 15min returns while $\sigma_t$ are realized
      volatilities calculated
      using 30s returns within each 15min interval. Horizontal axis:
      $z_t$}; Vertical axis: cummulative probability function (CDF) of
    $z_t$, arranged on such a scale that the CDF of the standard
    Gaussian is a straight line.}
  \label{fig:nordea_bank_15min_z_prob}
\end{figure}
% In addition, one can see
% from figure \ref{fig:nordea_15min_quotient_acf} and
% \ref{fig:nordea_15min_quotient_squared_acf} that there is essentially
% no auto-correlation in the $z_t$ or the $z_t^2$ series.
% \begin{figure}[htb!]
%   \centering
%   \subfigure[ACF of $z_t$]{
%     \includegraphics[scale=0.4, clip=true, trim=95 236 118
%     200]{../pics/nordea_15min_quotient_acf.pdf}
%     \label{fig:nordea_15min_quotient_acf}
%   }
%   \subfigure[ACF of $z_t^2$]{
%     \includegraphics[scale=0.4, clip=true, trim=95 236 118
%     200]{../pics/nordea_15min_quotient_squared_acf.pdf}
%     \label{fig:nordea_15min_quotient_squared_acf}
%   }
%   \caption{\small \it Nordea 15min $z_t$ and $z_t^2$ ACF.}
% \end{figure}

Andersen and Bollerslev et al reported that, for the exchange
rates between Deutch mark, yen and dollar, $\ln \sigma_t$ is
gaussian distributed \cite{Andersen03}. This is, however, not the case
for our series in question. In fact, in our case, $\ln \sigma_t$ is
right skewed (skewness 0.3342) and leptokurtic (kurtosis 
6.1006). See figure \ref{fig:nordea_15min_logvol_prob}
\begin{figure}[htb!]
  \centering
  %\vspace{-15mm}
  \includegraphics[scale=0.4, clip=true, trim=80 223 107
  4]{../pics/nordea_15min_logvol_prob.pdf}
  \caption{\small \it Probability plot of Nordea 15min
    $\ln\sigma_t$ unconditional distribution}
  \label{fig:nordea_15min_logvol_prob}
\end{figure}
Moreover, the series of $\ln\sigma_t$ shows long-lasting and
periodic autocorrelations with an apparent period of 33 (see figure
\ref{fig:nordea_15min_logvol_acf}). This suggests the series may be
described by a seasonal ARIMA model\footnote{ARIMA ---
  Integrated Autoregressive Moving Average} model. This we first
simplify the series by differencing \cite{BoxJenkins94}:
\[
w_t = (1-B)(1-B^s)\ln\sigma_t
\]
where $B$ is the back-shift operator and $s=33$ is the seasonality.

The autocorrelation function of the differenced process $w_t$, as
shown in figure \ref{fig:nordea_15min_w_acf}, clearly points to a seasonal
moving-average model: There are only 4 non-zero autocorrelations in
the plot, located at lags 1, 32, 33, 34, respectively; furthermore,
the two at 32 and 34 are approximately equal. Thus we can write down
the model as
\begin{eqnarray}
  w_t &=& (1 - \theta B)(1 - \Theta B^s) y_t \label{eq:nordea_w}
\end{eqnarray}
where $\theta$ and $\Theta$ are parameters to be determined and $y_t$
is a noise process with constant variance $\sigma_y^2$ and mean
0. $y_t$ is often refered to as the residuals.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of $\ln\sigma_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_logvol_acf.pdf}
    \label{fig:nordea_15min_logvol_acf}
  }
  \subfigure[ACF of $w_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_w_acf.pdf}
    \label{fig:nordea_15min_w_acf}
  }
  \caption{\small \it Nordea 15min $\ln\sigma_t$ and $w_t$
    autocorrelations}
  \label{fig:nordea1_15min_acf}
\end{figure}

@@

With an estimate of $\theta$ and $\Theta$, one can then infer the
noise process i.e. the residuals $y_t$:
\begin{equation}
  \label{eq:infer_y}
  y_t = w_t + \theta y_{t-1} + \Theta y_{t-s} - \theta \Theta y_{t-s-1}
\end{equation}
where we substitute $y_t\;(t \leq 0)$ with their unconditional
expectation 0.

In order to forecast the $w_t$ process, and hence the return process
itself, we must also know the distribution of $y_t$. Moreover, to
properly estimate the parameters of the model in the sense of maximum
likelihood, we are also in need of the distribution of $y_t$.

Figure \ref{fig:nordea_15min_y_qq} shows the normal probability plot of
$y_t$. It is evident from this figure that $y_t$ has fat tails.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.4, clip=true, trim=78 255 109
  123]{../pics/nordea2_y_normplot.pdf}
  \caption{\small \it Nordea 15min $y_t$ normal probability plot.}
  \label{fig:nordea_15min_y_qq}
\end{figure}
In addition, a simple calculation reveals that the distribution of
$y_t$ has skewness 0.2988 (shown in table
\ref{tab:nordea_15min_y_moments}).
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    mean & variance & skewness & kurtosis \\
    \hline
    0.0012 & 0.0935 & 0.2988 & 6.8691 \\
    \hline
  \end{tabular}
  \caption{\small \it Nordea 15min $y_t$ moments}
  \label{tab:nordea_15min_y_moments}
\end{table}
Based on this information, we find that $y_t$ can be well
described by a Johnson Su distribution \cite{Shang2004}:
\[
  y_t = \xi + \lambda\sinh{z_t - \gamma \over \delta}
\]
where $\gamma, \delta, \lambda, \xi$ are parameters to be determined
and $z_t \sim N(0, 1)$. The goodness of fitting is demonstrated in
figure \ref{fig:nordea_15min_y_js_fit} by the
empirical cummulative distribution function in comparison to the
theoretical one.
\begin{figure}[htb!]
  %\vspace{-18mm}
  \centering
    \includegraphics[scale=0.4, clip=true, trim=92 229 116
    133]{../pics/nordea_15min_y_js_fit.pdf}
    \caption{\small \it Nordea 15min residuals $y_t$ fitted to a
      Johnson Su distribution. Horizontal: values of $y_t$, denoted x;
      Vertical: $\ln\left(P(y_t < x)\right)$.}
    \label{fig:nordea_15min_y_js_fit}
\end{figure}
